---
title: "Lab 1, Stat 3503 Fall 2021"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: [css/xaringan-themer.css, css/extra.css]
    nature:
      highlightStyle: atom-one-light
      countIncrementalSlides: false
      ratio: 16:9
    includes:
      after_body: utils/ionicons.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, message=FALSE, warning=FALSE,
  fig.align="center", fig.retina=3
)

old_hooks <- fansi::set_knit_hooks(knitr::knit_hooks)

styles <- c(
  getOption('fansi.style'),  # default style
  "PRE.fansi CODE {background-color: transparent;}",
  "PRE.fansi-error {background-color: #DD5555;}",
  "PRE.fansi-warning {background-color: #DDDD55;}",
  "PRE.fansi-message {background-color: #EEEEEE;}"
)

old.hooks <- c(
  old_hooks,
  fansi::set_knit_hooks(
    knitr::knit_hooks,
    which=c('warning', 'error', 'message'),
    style=styles
  )
)

options(crayon.enabled=TRUE)
```

```{r clipboard, include=FALSE}
xaringanExtra::use_clipboard(
  button_text = "<ion-icon name=\"clipboard-outline\"></ion-icon>",
  success_text = "<ion-icon name=\"checkmark-outline\" style=\"color: #90BE6D\"></ion-icon>",
  error_text = "<ion-icon name=\"close-outline\" style=\"color: #F94144\"></ion-icon>"
)
```

```{r tile-view, include=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanthemer, eval=FALSE, include=FALSE}
library(xaringanthemer)

style_xaringan(
  text_color = "#f8f8f2",
  header_color = "#f8f8f2",
  background_color = "#282a36",
  link_color = "#ffb86c",
  text_slide_number_color = "#f8f8f2",
  code_inline_color = "#f8f8f2",
  code_inline_background_color = "#282c34",
  inverse_background_color = "#363428",
  inverse_text_color = "#f8f8f2",
  inverse_header_color = "#f8f8f2",
  text_font_google = google_font("IBM Plex Sans", "300", "300i"),
  header_font_google = google_font("IBM Plex Sans", "300", "300i"),
  code_font_google = google_font("IBM Plex Mono", "500"),
  outfile = "css/xaringan-themer.css"
)
```


# Lab 1 &mdash; September 20

## Question 1

### Load the tidyverse

```{r}
library(tidyverse)
theme_set(theme_bw())
```

- I am calling `theme_set(theme_bw())` so that my plots will have a white background, rather than
the default gray background
- See [here](https://ggplot2.tidyverse.org/reference/ggtheme.html#examples) for ggplot2's built-in
themes and previews of these themes

---

# Read in the data

```{r, message=TRUE}
weather <- read_csv("./data/ottawa-all-monthly.csv")
```

- Take note that `Month` was read in as a column of type character!

---

# Preview the data

```{r}
head(weather)
```

---

# Inspect the data

```{r}
dim(weather)
nrow(weather)
ncol(weather)
```

---
count: false

# Inspect the data

```{r}
names(weather)
length(names(weather))
```

---

# September

```{r}
September <- weather %>%
  filter(Month == "09")
```

---

# Plot 1

.pull-left[

```{r, eval=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()
```

]

---

# Plot 2

.pull-left[

```{r, eval=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r,echo=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

]

---

# Plot 3

.pull-left[

```{r, eval=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_line(aes(colour=Year))
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_line(aes(colour=Year))
```

]

---

# Plot 4

.pull-left[

```{r, eval=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_line(aes(colour=Year))+
  geom_smooth(method = "loess" ,formula = y ~ x)
```

- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
ggplot(September, aes(x=Date.Time, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_line(aes(colour=Year))+
  geom_smooth(method = "loess" ,formula = y ~ x)
```

]

---

# Question 2

## October

```{r}
October <- weather %>%
  filter(Month == "10")
```

---

# Combine September and October

```{r}
SeptOct <- bind_rows(September, October)

dim(SeptOct)
```

---

# Select columns

```{r}
SeptOct <- SeptOct %>%
  select(Year, Month, Mean.Max.Temp.C., Mean.Min.Temp.C., Total.Rain.mm., Total.Snow.cm.)
```

---

# Make new variable

```{r}
SeptOctnew <- SeptOct %>%
  mutate(Total.precip = Total.Rain.mm. + Total.Snow.cm. * 10)
```

---

# Plot 1

.pull-left[

```{r, eval=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()
```

]

---

# Plot 2

.pull-left[

```{r, eval=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

]

---

# Plot 3

.pull-left[

```{r, eval=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()+
  geom_line(aes(colour=Year))
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()+
  geom_line(aes(colour=Year))
```

]

---

# Plot 4

.pull-left[

```{r, eval=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()+
  geom_line(aes(colour=Year))+
  geom_smooth(method = "loess", formula = y ~ x)
```

- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip))+
  geom_point()+
  geom_line(aes(colour=Year))+
  geom_smooth(method = "loess", formula = y ~ x)
```

]

---

# New plot

.pull-left[

```{r, eval=FALSE}
ggplot(SeptOctnew, aes(x=Total.precip, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
ggplot(SeptOctnew, aes(x=Total.precip, y=Mean.Max.Temp.C.))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

]

---

# Fix that code

.pull-left[

```{r, eval=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip), colour=Month)+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

- Colour should be defined within the aesthetics
- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip), colour=Month)+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

]


---

# Fix that code

.pull-left[

```{r, eval=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip, colour=Month))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
ggplot(SeptOctnew, aes(x=Year, y=Total.precip, colour=Month))+
  geom_point()+
  geom_smooth(method = "loess", formula = y ~ x)
```

]

---

# Facetting

.pull-left[

```{r, eval=FALSE}
weather %>%
  filter(Month == "09" | Month == "10" | Month == "11") %>%
  mutate(Precipitation = Total.Rain.mm. + 25 * Total.Snow.cm.) %>%
  ggplot(aes(x=Year, y=Precipitation))+
  geom_line(aes(colour=Month))+
  geom_smooth(method = "loess", formula = y ~ x)+
  facet_wrap(~Month)
```

- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
weather %>%
  filter(Month == "09" | Month == "10" | Month == "11") %>%
  mutate(Precipitation = Total.Rain.mm. + 25 * Total.Snow.cm.) %>%
  ggplot(aes(x=Year, y=Precipitation))+
  geom_line(aes(colour=Month))+
  geom_smooth(method = "loess", formula = y ~ x)+
  facet_wrap(~Month)
```

]

---

# Better facetting

.pull-left[

```{r, eval=FALSE}
weather %>%
  filter(Month %in% c("09", "10", "11")) %>%
  mutate(Precipitation = Total.Rain.mm. + 25 * Total.Snow.cm.) %>%
  ggplot(aes(x=Year, y=Precipitation))+
  geom_line()+
  geom_smooth(method = "loess", formula = y ~ x)+
  facet_wrap(~Month, nrow=3, labeller="label_both")
```

- Can filter the months using less typing
- Don't need to colour by month anymore since we are facetting by month
- Clarify that the variable being facetted is `Month` using `labeller = "label_both"`, though
an even better option would be to use the full month names, e.g. September, October, November
- Change dimensions to 3x1 since time data is usually wide
- Specifying `method = "loess"` and `formula = y ~ x` are optional, but a warning will be shown
that this is what is used

]

.pull-right[

```{r, echo=FALSE}
weather %>%
  filter(Month %in% c("09", "10", "11")) %>%
  mutate(Precipitation = Total.Rain.mm. + 25 * Total.Snow.cm.) %>%
  ggplot(aes(x=Year, y=Precipitation))+
  geom_line()+
  geom_smooth(method = "loess", formula = y ~ x)+
  facet_wrap(~Month, nrow=3, labeller="label_both")
```

]
