---
title: "Lab 5, Stat 3503 Fall 2021"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: [css/extra.css, css/xaringan-themer.css]
    nature:
      highlightStyle: atom-one-light
      countIncrementalSlides: false
      highlightLines: true
      ratio: 16:9
    includes:
      after_body: utils/ionicons.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, message=FALSE, warning=FALSE,
  fig.align="center", dev="svglite"
)

old_hooks <- fansi::set_knit_hooks(knitr::knit_hooks)

styles <- c(
  getOption('fansi.style'),  # default style
  "PRE.fansi CODE {background-color: transparent;}",
  "PRE.fansi-error {background-color: #DD5555;}",
  "PRE.fansi-warning {background-color: #DDDD55;}",
  "PRE.fansi-message {background-color: #EEEEEE;}"
)

old.hooks <- c(
  old_hooks,
  fansi::set_knit_hooks(
    knitr::knit_hooks,
    which=c('warning', 'error', 'message'),
    style=styles
  )
)

options(crayon.enabled=TRUE)
```

```{r clipboard, include=FALSE}
xaringanExtra::use_clipboard(
  button_text = "<ion-icon name=\"clipboard-outline\"></ion-icon>",
  success_text = "<ion-icon name=\"checkmark-outline\" style=\"color: #90BE6D\"></ion-icon>",
  error_text = "<ion-icon name=\"close-outline\" style=\"color: #F94144\"></ion-icon>"
)
```

```{r tile-view, include=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanthemer, eval=FALSE, include=FALSE}
library(xaringanthemer)

style_xaringan(
  text_color = "#f8f8f2",
  header_color = "#f8f8f2",
  background_color = "#282a36",
  link_color = "#ffb86c",
  text_slide_number_color = "#f8f8f2",
  code_inline_color = "#f8f8f2",
  code_inline_background_color = "#282c34",
  inverse_background_color = "#363428",
  inverse_text_color = "#f8f8f2",
  inverse_header_color = "#f8f8f2",
  text_font_size = "0.9rem",
  header_h1_font_size = "1.75rem",
  header_h2_font_size = "1.50rem",
  header_h3_font_size = "1.25rem",
  text_font_google = google_font("IBM Plex Sans", "300", "300i"),
  header_font_google = google_font("IBM Plex Sans", "300", "300i"),
  code_font_google = google_font("IBM Plex Mono", "500"),
  code_font_size = "0.6rem",
  outfile = "css/xaringan-themer.css"
)
```

# Lab 5 &mdash; November 1

```{r, message=TRUE}
library(tidyverse)
library(broom)
```

---

# Reusing the data from last time

```{r}
bears <- read_csv("./data/bears.csv") %>%
  rename(active_hunters = `Active Hunters`) %>%
  filter(WMU != "Total") %>%
  filter(WMU %in% c("82A","83","84","73","74","75","76"))

bears
```

---

# Reusing the same linear model from last time

.pull-left[

```{r}
bears_lm <- lm(Harvest ~ active_hunters, data=bears)

summary(bears_lm)
```

]

---

# Creating a 95% confidence interval for the mean response

We are interested in a 95% confidence interval for the mean harvest when the number of active
hunters equals 150. Using the functions built into R, we can easily obtain this interval.

```{r}
predict(
  bears_lm,
  newdata = data.frame(active_hunters=150),
  interval = "confidence"
)
```

---

# Creating the interval manually

The formula for a 95% confidence interval for the mean response is:

$$\widehat{Y} \,\pm\, t_{1-\frac{\alpha}{2}}\,\widehat{\sigma}\sqrt{\frac{1}{n} \,+\, \frac{(X_{*} - \overline{X})^{2}}{\sum_{i}(X_{i} - \overline{X})^{2}}}$$

Let's collect all these values and store them in individual columns in a tibble.

---

# More on broom

- We can collected the required values by starting from the output of `broom::augment()` and
`broom::glance()`
- `broom::augment()` augments a model by returning the data used to fit the model and diagnostic
information

```{r, R.options=list(width=300)}
augment(bears_lm)
```

--

- `broom::glance()` returns general information on the model and goodness of fit measures

```{r, R.options=list(width=300)}
glance(bears_lm)
```

---

# Collecting the values

```{r}
augment(bears_lm) %>%
  summarise(
    xbar = mean(active_hunters),
    Sxx = sum((active_hunters - xbar)^2)
  ) %>%
  bind_cols(
    glance(bears_lm) %>%
      select(sigma, df.residual, nobs)
  )
```

---

# Collecting the values

```{r, R.options=list(width=300)}
augment(bears_lm) %>%
  summarise(
    xbar = mean(active_hunters),
    Sxx = sum((active_hunters - xbar)^2)
  ) %>%
  bind_cols(
    glance(bears_lm) %>%
      select(sigma, df.residual, nobs)
  ) %>%
  mutate(
    alpha = 0.05,
    beta0 = coef(bears_lm)[1],
    beta1 = coef(bears_lm)[2],
    xstar = 150,
    yhat = beta0 + beta1 * xstar,
    tval = qt(1 - alpha/2, df=df.residual)
  )
```

---

# Collecting the values

Finally, create a variable to store these values.

```{r, R.options=list(width=300)}
values <- augment(bears_lm) %>%
  summarise(
    xbar = mean(active_hunters),
    Sxx = sum((active_hunters - xbar)^2)
  ) %>%
  bind_cols(
    glance(bears_lm) %>%
      select(sigma, df.residual, nobs)
  ) %>%
  mutate(
    alpha = 0.05,
    beta0 = coef(bears_lm)[1],
    beta1 = coef(bears_lm)[2],
    xstar = 150,
    yhat = beta0 + beta1 * xstar,
    tval = qt(1 - alpha/2, df=df.residual)
  )
```

---

# Calculating the interval

```{r}
values %>%
  summarise(
    fit = beta0 + beta1 * xstar,
    lower = fit - tval * sigma * sqrt((1 / nobs) + (xstar - xbar)^2 / Sxx),
    upper = fit + tval * sigma * sqrt((1 / nobs) + (xstar - xbar)^2 / Sxx)
  ) %>%
  as.matrix() # only to see more digits
```

Compare them to our results from `predict()`:

```{r}
predict(
  bears_lm,
  newdata = data.frame(active_hunters=150),
  interval = "confidence"
)
```

---

# Creating a 90% prediction interval for response value of a new observation

We are interested in a 90% prediction interval for the response value of a new observation, where
the number of active hunters equals 55. Using the functions built into R, we can easily obtain this
interval.

```{r}
predict(
  bears_lm,
  newdata = data.frame(active_hunters=55),
  interval = "prediction",
  level = 0.90
)
```

---

# Creating the interval manually

The formula for a 90% confidence interval for the response value of a new observation is:

$$\widehat{Y} \,\pm\, t_{1-\frac{\alpha}{2}}\,\widehat{\sigma}\sqrt{1 + \frac{1}{n} \,+\, \frac{(X_{*} - \overline{X})^{2}}{\sum_{i}(X_{i} - \overline{X})^{2}}}$$

We can reuse the values that we had previously collected and update:

- `alpha`
- `xstar`
- `tval`

---

# Updating our values

```{r, R.options=list(width=300)}
values2 <- values %>%
  mutate(
    alpha = 0.10,
    xstar = 55,
    tval = qt(1 - alpha/2, df=df.residual)
  )

values2
```

---

# Calculating the interval

The formula differs only by a $1 +$ term under the square root.

```{r}
values2 %>%
  summarise(
    fit = beta0 + beta1 * xstar,
    lower = fit - tval * sigma * sqrt(1 + (1 / nobs) + (xstar - xbar)^2 / Sxx),
    upper = fit + tval * sigma * sqrt(1 + (1 / nobs) + (xstar - xbar)^2 / Sxx)
  ) %>%
  as.matrix() # only to see more digits
```

Compare them to our results from `predict()`:

```{r}
predict(
  bears_lm,
  newdata = data.frame(active_hunters=55),
  interval = "prediction",
  level = 0.90
)
```

---

# Extracting the design matrix

We can extract the design matrix, $X$, from our linear model using the `model.matrix()` function.

```{r}
X <- model.matrix(bears_lm)

head(X)
```

---

# Obtaining $X^{T}X$

```{r}
t(X) %*% X
```

---

# Obtaining the hat matrix

The hat matrix is defined as

$$H \,=\, X\,(X^{T}\,X)^{-1}\,X^{T}$$

```{r}
H <- X %*% solve(t(X) %*% X) %*% t(X)

H[1:6, 1:6]
```

---

# Obtaining the variance-covariance matrix of parameter estimates

The variance-covariance matrix of the parameter estimates is obtained as

$$\textbf{Var}(\widehat{\beta}) \,=\, \widehat{\sigma}^{2}(X^{T}\,X)^{-1}$$

We already have the value of $\widehat{\sigma}$ from our previous collection of values, so we just
need to remember to square it.

```{r}
var_beta <- values$sigma^2 * solve(t(X) %*% X)

var_beta
```

- The diagonal elements are the variances
- The off-diagonal elements are the covariances

---

# Standard errors

Since the variances were along the diagonal, if we square root them we will get the standard errors.
We can extract the diagonal elements of a matrix using `diag()`.

```{r}
se_beta <- sqrt(diag(var_beta))

se_beta
```

Compare with the standard errors computed from `summary()` or `broom::tidy()`:

```{r}
tidy(bears_lm)
```

They're the same (as they should be)!
