---
title: "Lab 3, Stat 3503 Fall 2021"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: [css/xaringan-themer.css, css/extra.css]
    nature:
      highlightStyle: null
      countIncrementalSlides: false
      ratio: 16:9
    includes:
      after_body: utils/ionicons.html
---

class: language-r
layout: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, message=FALSE, warning=FALSE,
  fig.align="center", dev="svglite"
)

old_hooks <- fansi::set_knit_hooks(
  knitr::knit_hooks,
  which = c("output", "message", "warning", "error")
)

options(crayon.enabled=TRUE)

htmltools::tagList(
  htmltools::htmlDependency(
    name = "prismr",
    version = "1.28.0",
    src = "libs/prismr",
    script = "prismr.js",
    stylesheet = "adam-one-light.css"
  )
)
```

```{r clipboard, include=FALSE}
xaringanExtra::use_clipboard(
  button_text = "<ion-icon name=\"clipboard-outline\"></ion-icon>",
  success_text = "<ion-icon name=\"checkmark-outline\" style=\"color: #90BE6D\"></ion-icon>",
  error_text = "<ion-icon name=\"close-outline\" style=\"color: #F94144\"></ion-icon>"
)
```

```{r tile-view, include=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanthemer, eval=FALSE, include=FALSE}
library(xaringanthemer)

style_xaringan(
  text_color = "#f8f8f2",
  header_color = "#f8f8f2",
  background_color = "#282a36",
  link_color = "#ffb86c",
  text_slide_number_color = "#f8f8f2",
  code_inline_color = "#f8f8f2",
  code_inline_background_color = "#282c34",
  inverse_background_color = "#363428",
  inverse_text_color = "#f8f8f2",
  inverse_header_color = "#f8f8f2",
  text_font_size = "0.9rem",
  header_h1_font_size = "1.75rem",
  header_h2_font_size = "1.50rem",
  header_h3_font_size = "1.25rem",
  text_font_google = google_font("IBM Plex Sans", "300", "300i"),
  header_font_google = google_font("IBM Plex Sans", "300", "300i"),
  code_font_google = google_font("IBM Plex Mono", "500"),
  code_font_size = "0.6rem",
  outfile = "css/xaringan-themer.css"
)
```

# Lab 3 &mdash; October 4

```{r, message=TRUE}
library(tidyverse)
library(lubridate)
```

---

# Soft-drink production data

- From 1950 to 1977, soft-drink production was reported on a quarterly basis
- From 1976 to 1995, soft-drink production was reported on a monthly basis
- We want to expand the quarterly production data by converting our monthly data into quarterly
data, and appending it to the end of the original quarterly data
- The final result should be a data set that ranges from 1950 to 1995 with production values being
reported each quarter

---

# Obtain quarterly soft-drinks data set

```{r, eval=FALSE}
if (!dir.exists("./data")) {
  dir.create("./data")
}

download.file("https://www150.statcan.gc.ca/n1/tbl/csv/16100100-eng.zip", destfile="./data/qdrinks.zip")
unzip("./data/qdrinks.zip", exdir="./data")
```

```{r, eval=FALSE}
file.rename("./data/16100100.csv", "./data/qdrinks.csv")
```

```{r, message=TRUE}
qdrinks <- read_csv("./data/qdrinks.csv")
```

---

# Inspect quarterly drinks data

```{r}
qdrinks %>%
  select(REF_DATE, VALUE)
```

- The `REF_DATE` column is in the format yyyy-mm
- We want to convert it to a full date format, yyyy-mm-dd
- The lubridate package contains functions to help us parse text into dates and datetimes
- Since our values are in the format yyyy-mm, we use the `ym()` function

---

# Modify quarterly drinks data

```{r}
quarterly <- qdrinks %>%
  select(REF_DATE, VALUE) %>%
  rename(date = REF_DATE, quarterly_value = VALUE) %>%
  mutate(date = ym(date))

quarterly
```

---

# Obtain monthly soft-drinks data set

```{r, eval=FALSE}
download.file("https://www150.statcan.gc.ca/n1/tbl/csv/16100099-eng.zip", destfile="./data/mdrinks.zip")
unzip("./data/mdrinks.zip", exdir="./data")
```

```{r, eval=FALSE}
file.rename("./data/16100099.csv", "./data/mdrinks.csv")
```

```{r, message=TRUE}
mdrinks <- read_csv("./data/mdrinks.csv")
```

---

# Inspect monthly drinks data

```{r}
mdrinks %>%
  select(REF_DATE, VALUE)
```

- The `REF_DATE` column is in the format yyyy-mm
- We want to convert it to a full date format, yyyy-mm-dd
- The lubridate package contains functions to help us parse text into dates and datetimes
- Since our values are in the format yyyy-mm, we use the `ym()` function

---

# Modify monthly drinks data

```{r}
monthly <- mdrinks %>%
  select(REF_DATE, VALUE) %>%
  rename(date = REF_DATE, monthly_value = VALUE) %>%
  mutate(date = ym(date))

monthly
```

---

# Convert monthly data to quarterly data

Starting with the `monthly` data set, create a new variable called `qdate` which classifies the
`date` by its corresponding quarter.

```{r}
monthly %>%
  mutate(qdate = quarter(date, with_year = TRUE))
```

---
count: false

# Convert monthly data to quarterly data

Group the rows by their resulting quarters and sum the monthly production values by group, resulting
in quarterly production values.

```{r}
monthly %>%
  mutate(qdate = quarter(date, with_year = TRUE)) %>%
  group_by(qdate) %>%
  summarise(quarterly_value = sum(monthly_value))
```

---
count: false

# Convert monthly data to quarterly data

```{r}
monthly %>%
  mutate(qdate = quarter(date, with_year = TRUE)) %>%
  group_by(qdate) %>%
  summarise(quarterly_value = sum(monthly_value))
```

We now have one row for each quarter for each year.

Note that after summarising, the last grouping is automatically dropped. Since we only grouped
on the values of one variable, the resulting data is fully ungrouped. Therefore, we do not need
to call `ungroup()`.

---
count: false

# Convert monthly data to quarterly data

Finally, we convert our quarterly dates (yyyy.q) back to full dates (yyyy-mm-dd).

```{r}
monthly_to_quarterly <- monthly %>%
  mutate(qdate = quarter(date, with_year = TRUE)) %>%
  group_by(qdate) %>%
  summarise(quarterly_value = sum(monthly_value)) %>%
  mutate(date = yq(qdate)) %>%
  select(date, quarterly_value)

monthly_to_quarterly
```

---

# Append the converted data to the quarterly data

Since we are appending our converted data to the end of the original quarterly data, we need to
make sure that the dates don't overlap (otherwise some dates will have multiple production values).

```{r}
quarterly_end_date <- quarterly %>%
  slice_max(date) %>%
  pull(date)

monthly_to_quarterly <- monthly_to_quarterly %>%
  filter(date > quarterly_end_date)
```

```{r}
full_quarterly <- bind_rows(quarterly, monthly_to_quarterly)

full_quarterly
```

---

# Check our work

Our data should range from 1950 to 1995 (46 years), with four quarters for each year. Therefore,
we should have 46 * 4 = 184 rows.

```{r}
full_quarterly %>%
  slice_min(date)

full_quarterly %>%
  slice_max(date)

full_quarterly %>%
  distinct(date) %>%
  nrow()
```

---

# Fit a simple linear regression model

We want to create a simple linear regression model with `quarterly_value` as the response and
`date` as the predictor.

```{r}
production <- lm(quarterly_value ~ date, data=full_quarterly)
```

---

# What are the estimated coefficients?

We can get the coefficients by using either:

```{r}
coef(production)
```

```{r}
summary(production)
```

---
count: false

# What are the estimated coefficients?

.pull-left[

```{r}
summary(production)
```

]

.pull-right[

- $\widehat{\beta}_{0} = 68275$
- $\widehat{\beta}_{1} = 7.541$

The equation of the fitted line is:

$$\widehat{y}_{i} \,=\, 68275 \,+\, 7.541x_{i}$$

]

---

# What are the standard errors of our estimates?

.pull-left[

```{r}
summary(production)
```

]

.pull-right[

- The standard error of $\widehat{\beta}_{0}$ is $1.089 \times 10^{3}$
- The standard error of $\widehat{\beta}_{1}$ is $0.219$

]

---

# What are the degrees of freedom of our model?

.pull-left[

```{r}
summary(production)
```

]

.pull-right[

- The degrees of freedom of the error is 182
- This is obtained by taking the number of observations in our data, $n$, minus the number of
non-intercept parameters estimated, $p$, minus 1, $n-p-1$.
- There are 184 data points and one non-intercept parameter estimated.

$$184 - 1 - 1 = 182$$

]

---

# Confidence intervals for our estimates

We can obtain confidence intervals for our estimates by wrapping our model with the `confint()`
function. By default, the confidence level is set to 95%.

```{r}
confint(production)
```

The 95% confidence interval for $\beta_{0}$ is $(66127,\, 70423)$.

The 95% confidence interval for $\beta_{1}$ is $(7.108,\, 7.974)$.
