---
title: "Lab 2, Stat 3503 Fall 2021"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: [css/xaringan-themer.css, css/extra.css]
    nature:
      highlightStyle: null
      countIncrementalSlides: false
      ratio: 16:9
    includes:
      after_body: utils/ionicons.html
---

class: language-r
layout: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, message=FALSE, warning=FALSE,
  fig.align="center", dev="svglite"
)

old_hooks <- fansi::set_knit_hooks(knitr::knit_hooks)

styles <- c(
  getOption('fansi.style'),  # default style
  "PRE.fansi CODE {background-color: transparent;}",
  "PRE.fansi-error {background-color: transparent;}",
  "PRE.fansi-warning {background-color: transparent;}",
  "PRE.fansi-message {background-color: transparent;}"
)

old.hooks <- c(
  old_hooks,
  fansi::set_knit_hooks(
    knitr::knit_hooks,
    which=c('warning', 'error', 'message'),
    style=styles
  )
)

options(crayon.enabled=TRUE)

htmltools::tagList(
  htmltools::htmlDependency(
    name = "prismr",
    version = "1.28.0",
    src = "libs/prismr",
    script = "prismr.js",
    stylesheet = "adam-one-light.css"
  )
)
```

```{r clipboard, include=FALSE}
xaringanExtra::use_clipboard(
  button_text = "<ion-icon name=\"clipboard-outline\"></ion-icon>",
  success_text = "<ion-icon name=\"checkmark-outline\" style=\"color: #90BE6D\"></ion-icon>",
  error_text = "<ion-icon name=\"close-outline\" style=\"color: #F94144\"></ion-icon>"
)
```

```{r tile-view, include=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanthemer, eval=FALSE, include=FALSE}
library(xaringanthemer)

style_xaringan(
  text_color = "#f8f8f2",
  header_color = "#f8f8f2",
  background_color = "#282a36",
  link_color = "#ffb86c",
  text_slide_number_color = "#f8f8f2",
  code_inline_color = "#f8f8f2",
  code_inline_background_color = "#282c34",
  inverse_background_color = "#363428",
  inverse_text_color = "#f8f8f2",
  inverse_header_color = "#f8f8f2",
  text_font_size = "0.9rem",
  header_h1_font_size = "1.75rem",
  header_h2_font_size = "1.50rem",
  header_h3_font_size = "1.25rem",
  text_font_google = google_font("IBM Plex Sans", "300", "300i"),
  header_font_google = google_font("IBM Plex Sans", "300", "300i"),
  code_font_google = google_font("IBM Plex Mono", "500"),
  code_font_size = "0.6rem",
  outfile = "css/xaringan-themer.css"
)
```

# Lab 2 &mdash; September 27

```{r, message=TRUE}
library(tidyverse)
library(gghighlight)
theme_set(theme_bw())
```

---

# Obtaining the flowers data set

```{r, eval=FALSE}
if (!dir.exists("./data")) {
  dir.create("./data")
}

download.file("https://www150.statcan.gc.ca/n1/tbl/csv/32100246-eng.zip", destfile="./data/flowers.zip")
unzip("./data/flowers.zip", exdir="./data")
```

```{r, eval=FALSE}
file.rename("./data/32100246.csv", "./data/flowers.csv")
file.rename("./data/32100246_MetaData.csv", "./data/flowers_metadata.csv")
```

```{r, message=TRUE}
flowers <- read_csv("./data/flowers.csv")
```

---

# Cleaning up

```{r}
flowers <- flowers %>%
  rename(year = REF_DATE, location = GEO, type = `Flowers and plants`)
```

- Although not asked, I am also renaming the `Flowers and plants` variable to `type`, because we
really shouldn't have spaces in variable names
- Variable names that violate the naming conventions can be referenced by wrapping them in backticks
as above

---

# Flower sales in Canada

.pull-left[

```{r, eval=FALSE}
can_flower_sales <- flowers %>%
  filter(location == "Canada", Output == "Sales")

ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()
```

]

.pull-right[

```{r, echo=FALSE}
can_flower_sales <- flowers %>%
  filter(location == "Canada", Output == "Sales")

ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()
```

]

---

# Why is this plot terrible?

.pull-left[

- Axis and legend names are not capitalised consistently: `VALUE`, `year`, `type`
- Legend is taking up too much plotting space
- Values in the legend contain unnecessary information
- Unclear what the y-axis, `VALUE`, is measuring
- Scientific notation on the y-axis requires the reader to do math

]

.pull-right[

```{r, echo=FALSE}
ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()
```

]

---

# Improve the plot &mdash; modify `type` values

We will modify the values in the `type` column of the original data set. Alternatively, you can
create a copy of the data set and modify the copy instead. The functions `str_remove` and 
`str_to_sentence` are found in the [stringr](https://stringr.tidyverse.org/) package. Finally, we
can re-obtain the required data for this question. 

.pull-left[

```{r}
flowers <- flowers %>%
  mutate(
    type = str_remove(type, pattern="Total\\s"),
    type = str_remove(type, pattern="\\s\\[.*\\]"),
    type = str_to_sentence(type)
  )
```

- Remove the phrase "Total" with the space that follows, from the `type` values
- Remove the space before the square brackets, the square brackets, and anything contained within,
from the `type` values
- Capitalise the first letter of each value

```{r}
can_flower_sales <- flowers %>%
  filter(location == "Canada", Output == "Sales")
```

]

--

.pull-right[

Check our modifications:

```{r}
can_flower_sales %>%
  distinct(type)
```

]

---

# Improve the plot &mdash; change scale of y-axis

To work around the scientific notation, change the unit of the y-axis from dollars to millions of
dollars.

```{r}
can_flower_sales <- can_flower_sales %>%
  mutate(VALUE = VALUE / 1e6)
```

---

# Improve the plot &mdash; make the new plot

.pull-left[

Starting with the base plot:

```{r, eval=FALSE}
ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()
```

]

---

# Improve the plot &mdash; make the new plot

.pull-left[

Label the lines on the plot:

```{r, eval=FALSE}
ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()+
  gghighlight(label_params = list(alpha=0.6))
```

- By default, `gghighlight::gghighlight` highlights and labels five objects
- Coincidentally, we have exactly five lines here
- Transparency is added to the label so that we can still see the line underneath

]

.pull-right[

```{r, echo=FALSE}
ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()+
  gghighlight(label_params = list(alpha=0.6))
```

]

---

# Improve the plot &mdash; make the new plot

.pull-left[

Fix axis labels and add titles

```{r, eval=FALSE}
ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()+
  gghighlight(label_params = list(alpha=0.6))+
  labs(
    x="Year", y="Total sales (million dollars)",
    title="Plant sales in Canada", subtitle="2007 to 2020"
  )
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(can_flower_sales, aes(x=year, y=VALUE, group=type, colour=type))+
  geom_line()+
  gghighlight(label_params = list(alpha=0.6))+
  labs(
    x="Year", y="Total sales (million dollars)",
    title="Plant sales in Canada", subtitle="2007 to 2020"
  )
```

]

---

# Flower production in Canada in 2020

.pull-left[

```{r, eval=FALSE}
can_flower_prod <- flowers %>%
  filter(year == 2020, location == "Canada", Output == "Production (number)")

ggplot(can_flower_prod, aes(x=type, y=VALUE))+
  geom_col()
```

- Note that `geom_col` is used when the counts already exist in your data (in this case, `VALUE`)
- `geom_bar` is used when you do not already have the counts; in this case, 
[ggplot2](https://ggplot2.tidyverse.org/) will perform the counting
- Recall that we had previously modified the values of the `type` variable to remove the extra
clutter

]

.pull-right[

```{r, echo=FALSE}
can_flower_prod <- flowers %>%
  filter(year == 2020, location == "Canada", Output == "Production (number)")

ggplot(can_flower_prod, aes(x=type, y=VALUE))+
  geom_col()
```

]

---

# How can we improve this plot?

.pull-left[

- Consistent capitalisation of axis names
- Scale `VALUE` so that there is no scientific notation
- Depending on the size of the output plot, the `type` labels may end up overlapping; flip the axes
so that this won't happen
- Notice that the `type` axis is sorted by alphabetical order; reorder the bars so that they appear
"in order"
- Give a more descriptive label for `VALUE` and add a title to the plot

]

.pull-right[

```{r, echo=FALSE}
can_flower_prod <- flowers %>%
  filter(year == 2020, location == "Canada", Output == "Production (number)")

ggplot(can_flower_prod, aes(x=type, y=VALUE))+
  geom_col()
```

]

---

# Improve the plot &mdash; make the new plot

.pull-left[

Flip the axes. Since [ggplot2](https://ggplot2.tidyverse.org/) v3.3.0+, we can just swap
the aesthetic variables instead of using `coord_flip`.

```{r, eval=FALSE}
ggplot(can_flower_prod, aes(x=VALUE, y=type))+
  geom_col()
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(can_flower_prod, aes(x=VALUE, y=type))+
  geom_col()
```

]

---

# Improve the plot &mdash; change scales and ordering

To work around the scientific notation, change the unit of `VALUE` from a single unit to millions of
units.

At the same time, we can reorder `type` by *ascending* value of `VALUE` using `fct_reorder` from the
[forcats](https://forcats.tidyverse.org/) package. In reordering the levels of `type`, we are
converting it from character (text) to an ordered factor (ordered categorical variable).

```{r}
can_flower_prod <- can_flower_prod %>%
  mutate(
    VALUE = VALUE / 1e6,
    type = fct_reorder(type, VALUE)
  )
```

- Since we are making a horizontal bar chart, we reorder `type` by *ascending* value of `VALUE`
- If we had decided to stick with a vertical bar chart, we would need to reorder `type` by
*descending* value of `VALUE` using:

```{r, eval=FALSE}
fct_reorder(type, VALUE, .desc=TRUE)
```

---

# Improve the plot &mdash; make the new plot

.pull-left[

Check our work by re-running the same code that produced the previous plot:

```{r, eval=FALSE}
ggplot(can_flower_prod, aes(x=VALUE, y=type))+
  geom_col()
```

]

.pull-right[

```{r, echo=FALSE}
ggplot(can_flower_prod, aes(x=VALUE, y=type))+
  geom_col()
```

]

---

# Improve the plot &mdash; make the new plot

.pull-left[

Fix the axis names and add a title.

```{r, eval=FALSE}
ggplot(can_flower_prod, aes(x=VALUE, y=type))+
  geom_col()+
  labs(
    x="Number produced (millions)", y="Plant type",
    title="Number of plants produced in Canada in 2020, by plant type"
  )
```

- You can add some colour to this plot, but I'm not going to

]

.pull-right[

```{r, echo=FALSE}
ggplot(can_flower_prod, aes(x=VALUE, y=type))+
  geom_col()+
  labs(
    x="Number produced (millions)", y="Plant type",
    title="Number of plants produced in Canada in 2020, by plant type"
  )
```

]
