---
title: "Lab 6, Stat 3503 Fall 2021"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: [css/extra.css, css/xaringan-themer.css]
    nature:
      highlightStyle: null
      countIncrementalSlides: false
      highlightLines: true
      ratio: 16:9
    includes:
      after_body: utils/ionicons.html
---

class: language-r
layout: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE, message=FALSE, warning=FALSE,
  fig.align="center", dev="svglite"
)

old_hooks <- fansi::set_knit_hooks(
  knitr::knit_hooks,
  which = c("output", "message", "warning", "error")
)

options(crayon.enabled=TRUE)

htmltools::tagList(
  htmltools::htmlDependency(
    name = "prismr",
    version = "1.28.0",
    src = "libs/prismr",
    script = "prismr.js",
    stylesheet = "adam-one-light.css"
  )
)
```

```{r clipboard, include=FALSE}
xaringanExtra::use_clipboard(
  button_text = "<ion-icon name=\"clipboard-outline\"></ion-icon>",
  success_text = "<ion-icon name=\"checkmark-outline\" style=\"color: #90BE6D\"></ion-icon>",
  error_text = "<ion-icon name=\"close-outline\" style=\"color: #F94144\"></ion-icon>"
)
```

```{r tile-view, include=FALSE}
xaringanExtra::use_tile_view()
```

```{r xaringanthemer, eval=FALSE, include=FALSE}
library(xaringanthemer)

style_xaringan(
  text_color = "#f8f8f2",
  header_color = "#f8f8f2",
  background_color = "#282a36",
  link_color = "#ffb86c",
  text_slide_number_color = "#f8f8f2",
  code_inline_color = "#f8f8f2",
  code_inline_background_color = "#282c34",
  inverse_background_color = "#363428",
  inverse_text_color = "#f8f8f2",
  inverse_header_color = "#f8f8f2",
  text_font_size = "0.9rem",
  header_h1_font_size = "1.75rem",
  header_h2_font_size = "1.50rem",
  header_h3_font_size = "1.25rem",
  text_font_google = google_font("IBM Plex Sans", "300", "300i"),
  header_font_google = google_font("IBM Plex Sans", "300", "300i"),
  code_font_google = google_font("IBM Plex Mono", "500"),
  code_font_size = "0.6rem",
  outfile = "css/xaringan-themer.css"
)
```

# Lab 6 &mdash; November 8

```{r, message=TRUE}
library(tidyverse)
library(broom)
theme_set(theme_bw())
```

---

# Diamonds data set

```{r}
data(diamonds)

diamonds
```

---

# Create a variable for the volume

```{r}
diamonds <- diamonds %>%
  mutate(volume = x * y * z)

diamonds
```

---

# Remove zero-volume diamonds

```{r}
diamonds <- diamonds %>%
  filter(volume > 0)
```

---

# One diamond has a suspiciously high volume

There is one observation with an extremely large volume.

```{r}
diamonds %>%
  arrange(desc(volume))
```

---

# One diamond has a suspiciously high volume

There is one observation with an extremely large volume.

We can extract this row using:

```{r}
diamonds %>%
  slice_max(volume)
```

---

# One diamond has a suspiciously high volume

We can extract the volume value using:

```{r}
diamonds %>%
  slice_max(volume) %>%
  pull(volume)
```

We don't actually need the exact value to remove it from the data set. We can just use an arbitrary
cutoff such as 3000.

---

# Remove this diamond from the data

To remove this observation, we can use:

```{r}
diamonds <- diamonds %>%
  filter(volume < 3000)
```

Check our work:

```{r}
diamonds %>%
  arrange(desc(volume))
```

---

# Make a scatterplot

.pull-left[

We wish to make a linear model to predict carat using volume. Let us check that there exists a
linear relationship between the two variables.

```{r, eval=FALSE}
ggplot(diamonds, aes(x=volume, y=carat))+
  geom_point()
```

There are actually .red[**two**] points with identical carat and volume values where the volume is
abnormally large for such low carat values. (See output from previous slide).

For now, we will ignore these points.

]

.pull-right[

```{r, echo=FALSE}
ggplot(diamonds, aes(x=volume, y=carat))+
  geom_point()
```

]

---

# Fit the model

```{r}
carat_lm <- lm(carat ~ volume, data=diamonds)

summary(carat_lm)
```

The equation of the fitted line is:

$$\widehat{Y}_{i} \,=\, -3.324*10^{-3} \,+\, 6.170*10^{-3}X_{i}$$

---

# Create 80% confidence intervals for the parameter estimates

Produce 80% interval estimates for the intercept and effect of 1 unit change in diamond volume, i.e.
confidence intervals for $\beta_{0}$ and $\beta_{1}$.

```{r}
confint(carat_lm, level=0.80)
```

---

# Hypothesis testing

Test the one sided null hypothesis that the slope is less than 0.005.

$$H_{0}: \beta_{1} \leq 0.005, \quad H_{A}: \beta_{1} > 0.005$$

The value of the test statistic is computed as:

$$T \,=\, \frac{6.170*10^{-3} \,-\, 5*10^{-3}}{2.015*10^{-6}}$$

```{r}
tstat <- carat_lm %>%
  tidy() %>%
  filter(term == "volume") %>%
  summarise(tstat = (estimate - 0.005) / (std.error)) %>%
  pull(tstat)

tstat
```

```{r}
tstat_df <- glance(carat_lm) %>%
  pull(df.residual)

tstat_df
```


---

# Hypothesis testing

Since this is an upper-tailed test, the $p$-value is the area to the right of `tstat`.

```{r}
pt(tstat, df=tstat_df, lower.tail=FALSE)
```

Since the $p$-value is less than $\alpha = 0.05$, we reject the null hypothesis and conclude that
the true value of the slope parameter is greater than 0.005.

---

# Test for model usefulness

Test the null hypothesis that the regression model in (b) is not relevant.

$$H_{0}: \beta_{1} \,=\, 0, \quad H_{A}: \beta_{1} \neq 0$$

For the one parameter case, this test can be carried out via $t$-test.

```{r}
tidy(carat_lm) %>%
  filter(term == "volume")
```

The $p$-value of this test is zero. Since the $p$-value is less than $\alpha=0.05$, we can reject
the null hypotheis and conclude that our model is useful.

---

# Test for model usefulness

Test the null hypothesis that the regression model in (b) is not relevant.

$$H_{0}: \beta_{1} \,=\, 0, \quad H_{A}: \beta_{1} \neq 0$$

For one or parameters, this test can be carried out via $F$-test.

```{r}
anova(carat_lm) %>%
  tidy()
```

The $p$-value is zero so we can once again reject the null hypothesis and conclude that our model is
useful.

---

# Test for model usefulness

Test the null hypothesis that the regression model in (b) is not relevant.

$$H_{0}: \beta_{1} \,=\, 0, \quad H_{A}: \beta_{1} \neq 0$$

We could have actually read the results of the $F$-test from the model summary.

```{r, eval=FALSE}
## Call:
## lm(formula = carat ~ volume, data = diamonds)
## 
## Residuals:
##     Min      1Q  Median      3Q     Max 
## -4.6600 -0.0081 -0.0016  0.0056  1.0073 
## 
## Coefficients:
##               Estimate Std. Error t value Pr(>|t|)    
## (Intercept) -3.324e-03  3.037e-04  -10.95   <2e-16 ***
## volume       6.170e-03  2.015e-06 3062.32   <2e-16 ***
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
## 
## Residual standard error: 0.03582 on 53917 degrees of freedom
## Multiple R-squared:  0.9943,	Adjusted R-squared:  0.9943 
## F-statistic: 9.378e+06 on 1 and 53917 DF,  p-value: < 2.2e-16 #<<
```

---

# Test for model usefulness

Test the null hypothesis that the regression model in (b) is not relevant.

$$H_{0}: \beta_{1} \,=\, 0, \quad H_{A}: \beta_{1} \neq 0$$

Or using `broom::glance()`:

```{r}
glance(carat_lm) %>%
  select(statistic, df, df.residual, p.value) %>%
  rename(
    fvalue = statistic,
    df1 = df,
    df2 = df.residual
  )
```
